---
title: "2 Calculation Exercises"
author: "Andrew Boschee"
date: "6/5/2018"
output:
  word_document: default
  html_document: default
---

```{r,echo=FALSE}
options(digits=12)
unit.test.points=0
```

# General instructions.

Exercise 1 is required and must be solved using both R and SAS. This exercise is worht 20 points, 10 points each for R and SAS. The R portion is included in this document and includes a unit test that checks the results of your calculations. 

Exercises 2 is also required, but you may choose to use either R or SAS. Make that you include in your code comments documeting your choice, don't simply submit the results without explanation. This exercise is worth 10 points.

Exercises 3-5 are optional, choose two of three and document your choice. Your are also allowed to choose from either R or SAS. Choose only 1, each will be worth 10 points, and the entire assignment will be worth 50 points.

Be sure to document your code and comment on the results.

# Exercise 1 (R and SAS, required)


Calculate Cohen's $d$ for calories per recipe, calories per serving and servings per recipe, comparing years 1936 and 2006. Use the formula
$$
d = \frac{|m_1-m_2|}{\sqrt{(s_1^2 + s_2^2)/2}}
$$

## Part a. 
Enter the R code in the code chunk below, and assign a value to `d_12` for unit testing.

Declared variables for the mean and standard deviation in 1936 and 2006 for calories per serving, servings per recipe, and calories per recipe. Calories per recipe is assigned to d_12 for unit testing.


```{r}
s_i <- 1050.0
m_i <- 2123.8
sd_1936_cps <- 124.8
sd_2006_cps <- 168.3
m_1936_spr <- 12.9
m_2006_spr <- 12.7
s_1936_spr <- 13.3
s_2006_spr <- 13.0
cps_1936 <- 268.1
cps_2006 <- 436.9
m_j <- 3051.9
s_j <-1496.2

d_calories_per_serving <- (abs(cps_2006-cps_1936))/(sqrt((sd_1936_cps^2 + sd_2006_cps^2)/2))
d_calories_per_serving
d_servings_per_recipe <- (abs(m_1936_spr-m_2006_spr))/(sqrt((s_1936_spr^2 + s_2006_spr^2)/2))

d_12 <- (abs(m_i-m_j))/(sqrt((s_i^2 + s_j^2)/2))

```

```{r, unit test for exercise 1,echo=FALSE}
if(abs(d_12-0.7181)<0.0001) {
  print(unit.test.points <- unit.test.points+8)
} else {
  cat('d is not assigned the correct value','\n')
}
```

## Part b.
For the SAS portion, you may use PROC IML or macro language. Use `2 Calculations Template.sas` to start. This file contains unit tests for Exercises 1 and 2.

# Exercise 2 (R or SAS, required)

The probablity of an observation $x$, when taken from a normal population with mean $\mu$ and variance $\sigma^2$ is calculated by
$$
L (x ; \mu, \sigma^2) = \frac{1}{\sigma \sqrt{2 \pi}^{}} e^{- \frac{(x - \mu)^2}{2 \sigma^2}}
$$
For values of $x = \{ 0.1, 0.2 \}$, write code to calculate $L (x ; \mu = 0, \sigma = 1)$. Assign the values to `l_1` and `l_2`.
  
### Answer

If you choose to do this exercise in R, use the unit test below.

```{r}
mu <- 0
x_1 <- 0.1
x_2 <- 0.2
variance_1 <- 1


x_minus_mean_sq_1 <- -1*((x_1-mu)^2)
x_minus_mean_sq_2 <- -1*((x_2-mu)^2)

sqrt_two_pi <- sqrt(2*pi)
1/(sqrt_two_pi)*exp(1*(-1*((x_1-mu)^2))/(2*variance_1))


l_1 <- 1/(sqrt_two_pi)*exp(1*(x_minus_mean_sq_1)/(2*variance_1))

l_2 <- 1/(sqrt_two_pi)*exp(1*(x_minus_mean_sq_2)/(2*variance_1))
```



### Unit Test (R)
```{r, unit test for exercise 2,echo=FALSE}
if(abs(l_1 - 0.396952)<1e-6) {
  print(unit.test.points <- unit.test.points + 4)
} else {
  print("likelihood for x = 0.1 incorrect")
}

if(abs(l_2 - 0.391043)<1e-6 ) {
  print(unit.test.points <- unit.test.points + 4)
} else {
  print("likelihood for x = 0.2 incorrect")
}
```

If you choose to do this exercise in SAS, you must do this in PROC IML, using the unit test template `2 Calculations.sas`.

# Exercise 3 (R or SAS)

## Part a

Write code to compute
  
  $$7 - 1 \times 0 + 3 \div 3$$
  
Type this in verbatim, using only numbers, `-`,`*` and `/`, with no parenthesis. Do you agree with the result? Explain why, one or two sentences.

#### Answer
I agree with the answer due to the order of operations. Multiplication and division will be done first going left to right and followed by addition and subtraction resulting in 8
```{r}
7-1*0+3/3
```

## Part b

  According to "Why Did 74% of Facebook Users Get This Wrong?" (<http://www.classroomprofessor.com/teaching-math/why-did-74pc-of-facebook-users-get-this-wrong/>), most people would compute the result as 1. 
  Use parenthesis `( )` to produce this result.

#### Answer

```{r}
(7-1)*0+(3/3)
```

## Part c

Several respondents to the survey cited in Part 2 gave the answer 6. Add *one* set of parenthesis to produce this result.

#### Answer
```{r}
7-1*(0+3/3)
```

# Exercise 4. (R or SAS)

### Part a

Quoting from Wansink and Payne

> Because of changes in ingredients,
the mean average calories in a recipe increased by 928.1 (from
2123.8 calories ... to 3051.9 calories
... ), representing a 43.7% increase.

Show how 43.7% is calculated from 2123.8 and 3051.9, and confirm W&P result.

> The resulting increase of 168.8 calories (from 268.1 calories ... to 436.9 calories ...) represents a 63.0% increase ... in calories per serving.

### Part b

Repeat the calculations from above and confirm the reported 63.0% increase in calories per serving.

Part A shown with declared variable 'cprchange' to define average calories in a recipe change resulting in 43.7%.
Part B shown with 'cpschange' to define the change in average calories per serving resulting in 63% increase.

```{r}
cprchange <- (m_j - m_i)/(m_i)
print(cprchange)



cpschange <- (cps_2006 - cps_1936)/cps_1936
print(cpschange)
```

### Part c

Using values from Table 1 Wansink, calculate the percent change in calories per serving from 1997 to 2006. How does this value compare to 63.0 increase as quoted in part b.?

Calculation shown below with new variable declared using calories per serving from wansink table


```{r}
table_cps_2006 <- 384.4
cps_1997 <- 288.6

cpschange_97_to_06 <- (table_cps_2006 - cps_1997)/cps_1997
print(cpschange_97_to_06)

```


# Exercise 5. (R or SAS)

From Wansink and Payne
> Given that the average 2006 recipe had 1.1 fewer servings than in 1936, the average calorie density per serving size has increased by 37.4% 

From https://foodpsychology.cornell.edu/research/joy-cooking-too-much-70-years-calorie-increases-classic-recipes
> "Also, the mean average calories per serving increased in 17 out of the 18 recipes by 37.4% from 268.1 calories to 436.9 calories"

https://foodpsychology.cornell.edu/discoveries/joy-cooking-too-much
>This expanded portion size helps explain why calories per serving have increased from an average of 168.8 calories to 436.9 calories, which is a 63% increase in calories per serving

Show calculations in R or SAS to either confirm these statements (using values from Wansink Table 1) or show where the quoted values are inconsistent with Wansink Table 1.


# Total points from unit tests
```{r}
unit.test.points
```



